OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-11-18 manuscript_nar.tex
First draft after first round of revisions.

\address{%
$^{1}$Gene Regulation, Stem Cells and Cancer Program, Centre for
Genomic Regulation (CRG), The Barcelona Institute of Science and
Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain
and
$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain
and
$^{3}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
Barcelona, Spain
and
$^{4}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 1 kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Neither of these strategies were designed for cell types with karyotypic
aberrations, most common in cancer. Yet, Hi-C is very sensitive to
aneuploidy, copy number variations and translocations.  Actually, these
aberrations have so much influence on the outcome that they can be used as
signatureto re-assemble the target genome \citep{korbel2013genome}. An
additional complication is that karyotypic aberrations are not
experimental biases, so it is unclear whether they should be corrected at
all or be considered part of the biological signal.

@@

\subsection{Model}
\label{sec:model}

@@

\subsection{Copy number correction}
\label{sec:hmm}

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
($\lambda_i'$). The model consists of 7 states that correspond to 1, 2,
3, 4, 5, 6 and 8 copies of the target, for a total of 3 emission
parameters (a single scaling parameter, a single standard deviation and a
single degree of freedom for all the states) and 21 transition parameters.

@@

To test the correction of biases, we gathered a set of
published \citep{ledily2014distinct, encode2012integrated, rao20143d,
stadhouders2017transcription, lin2012global, dixon2012topological} and
unpublished Hi-C data of different cell types and organisms
(Table \ref{tab:samples}).

@@

%\begin{landscape}

\begin{table*}
\label{tab:samples}
{\begin{tabular}{ccccccc}
  \hline
  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
  \textbf{Source} \\
  \hline
dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D & NA \\
dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
HindIII\_T0 & T47D & dilution Hi-C &
  HindIII & CRG & T47D & SRR1054341 \\
NcoI\_T0    & T47D & dilution Hi-C &
  NcoI      & CRG & T47D & SRR1054343 \\
ENCLB758KFU & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB758KFU \\
ENCLB183QHG & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB183QHG \\
HIC069  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658693 \\
HIC070  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658694 \\
HIC071  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
HIC074  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
GSM987817 & B cell  & dilution Hi-C &
  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
GSM987818 & B cell  & dilution Hi-C &
  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
GSM862720 & ES cell & dilution Hi-C &
  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
GSM862721 & ES cell & dilution Hi-C &
  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
GSM862722 & ES cell & dilution Hi-C &
  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
  \hline
\end{tabular}}{}
\caption{Hi-C data sets used in this study. Sample ID: Unique sample
identifier in the respective projects. Cell type: source of the biological
sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
digestion. Sequencing core: Laboratory performing the experiment (CRG:
Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
Baylor College of Medicine; UCSD: University of California San Diego;
LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
data identifier.}
\end{table*}

%\end{landscape}

@@

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 and mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks}. For more details, see the methods section of
\cite{stadhouders2017transcription}. In addition, the pipeline is
available from the supplementary material published by
\cite{quilez2017managing}.

We developed the routines contained in the \texttt{dryhic} R package (available
at http://www.github.com/qenvio/dryhic) to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences at different
resolutions (not shown).

\subsection{Comparison of Hi-C matrices}
\label{sec:comp}

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the last option, proposed by
\cite{yan2017hicspector} is to measure the Pearson correlation between the
last eigenvectors of the Laplacian of the Hi-C matrix. This approach
borrows the concepts of spectral clustering \citep{von2007tutorial} and
amounts to comparing high level features of the matrix.

We defined three data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples (see Table
\ref{tab:samples} for details). Given a set of experiments and a metric,
we first computed all pairwise combinations between experiments and then
classified the comparisons according to the characteristics of each pair
(cell type, protocol, batch and treatment).

@@

\subsection{Experimental bias correction}

The principle of \textit{OneD} is to explicitly model Hi-C biases on a
single variable: the total amount of contacts for each bin of the matrix
(see \ref{sec:model} for detail). The reason for this choice is that the
total amount of contacts is approximately proportional to the local copy
number. For instance, a duplicated region in a diploid genome will show on
average a 50\% increase in the number of contacts. Discontinuities of the
amount of contacts thus correspond to changes of the copy number.

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]{img/figure1.pdf}}
\caption{Total amount of contacts per bin. A. Non linear relationship
between the number of restriction sites and the total number of
contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The brown line represents the raw
signal, the blue line represents the signal after bias correction, the
black line represents the signal after bias and copy number (CN) correction.
The long arm of chromosome 17 (the region corresponding to 20-80 Mbp) is
present in four copies, explaining that the signal is about twice higher
than for the short arm.}
\label{fig:totals}
\end{figure}

Experimental biases affect the total amount of contacts in a continuous
but not necessarily linear way. Figure~\ref{fig:totals}A shows the
relationship between the amount of contacts and the number of restriction
enzyme sites in T47D, a breast cancer cell line with an aberrant
karyotype.  Four clouds are visible. Each corresponds to sequences present
in one to four copies. In all of them, the relationship flattens as the
number of restriction sites increases. To capture this
relationship, \textit{OneD} fits a non-linear model between the total
amount of contacts and the known sources of bias (by default the G+C
content, the number of restriction sites and the mappability of the
reads).

The experimental biases are estimated genome-wide and each cell of the
matrix is then corrected using equation (\ref{eq:xhat}). Note that the
corrected amount of contacts is still proportional to the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
total amount of contacts (blue line).

% Marc's comment here was regarding lack of motivation and explanation for
% the copy number comparison. I've rescued some text from an older
% version. We should also comment here about CN being a bias or providing
% biological information, so to correct or not to correct for CN

In what follows, we benchmarked \textit{OneD}, against \textit{vanilla},
\textit{ICE} \citep{imakaev2012iterative}, \textit{caICB}
\citep{wu2016computational} and the Local Genomic Features method
\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
methods correct biases implicitly, whereas the fourth method does it
explicitly.

Given that our model is based on total number of contacts, we reasoned
that a preliminary test would be to check if the corrected number of
contacts per bin reflects the copy number (as measured by an independent
technique such as array-based copy-number segmentation). We tested the
validity of this approach against the Catalogue Of Somatic Mutations In
Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
shows the Pearson correlation between the corrected number of contacts and
the copy number estimation for T47D and K562 (a leukemia cell line with an
aberrant karyotype). Similar results were obtained using Spearman
correlation (Supplementary Figure 1).  All the methods except
\textit{OneD} decreased the agreement between the signal and the copy
number because they partially corrected the discrepancy. In contrast,
\textit{OneD} enhanced the conformity of the signal with the copy number.
Not correcting for variable copy number at that stage may seem
counter-intuitive, but the tests below will show this leads to better
performance.

\begin{figure}
	\centerline{\includegraphics[width=.39\textwidth]
{img/copy_number_figure2.pdf}}
\caption{Pearson correlation between the total number of contacts per bin
and an independent estimation of the copy number (COSMIC). Left panel
T47D breast cancer cell line, right panel K562 leukemia cell line.}
\label{fig:copy_number}
\end{figure}

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variance, such as batch effects, laboratory of
origin and protocol variations. Similarly, a good normalization should
decrease the similarity between different samples to enhance the
biological differences.

We assembled two Hi-C data sets obtained from T47D and K562 cells.  In
each set we spiked two samples from the other cell line
(Table~\ref{tab:samples}) to introduce biological variability. We compared
matrices before and after normalization by different methods using the
Pearson correlation of observed over expected counts (see \ref{sec:comp}).
This gave an indication of the impact of a given normalization method. The
results are summarized in Figure~\ref{fig:aberrant}.

The \textit{caICB} and \textit{ICE} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an
undesirable effect, as it obscures the biological variability. Likewise,
these methods decreased the similarity between samples that received the
same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
normalization process is detrimental to the biological signal in these two
cases. The method \textit{vanilla} followed the same trend but to a lesser
extent, consistent with the fact that it consists of a single \textit{ICE}
iteration. \textit{OneD} was the only method to increase the similarity
between experiments carried out on the same material but with a different
protocol (Figure~\ref{fig:aberrant}A).

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). All the methods, including the
absence of normalization, succeeded in identifying the T47D outliers among
the K562 samples, but recognizing the K562 outliers among the T47D samples
proved more challenging. \textit{OneD} increased the discrimination power
compared to the raw matrices, but all the other methods decreased it.
Actually, they performed little better than random on this task. Using the
other metrics described in \ref{sec:comp} yielded similar results
(Supplementary Figures 5, 6 and 7). Note that Spearman correlation of
contacts presented the worst performance for all scenarios, and it thus
seems to be a poor choice for a metric to compare Hi-C matrices.

Taken together, these results show that correcting experimental biases
with \textit{OneD} enhances the biological variation and reduces the noise
on samples with an aberrant karyotype.

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotypes. A and
B. Average changes compared to the raw data on the T47D and K562 sets. The
bars represent 95\% confidence intervals centered on the mean difference
of the correlation score between a given correction method and the raw
data. C and D. ROC curves on the T47D and K562 sets.  The areas under the
curve are indicated in the bottom right corner. The color code is the same
as panels A and B. The brown lines correspond to raw matrices. All results
in this figure are based on Pearson correlations between the observed over
expected counts.}
\label{fig:aberrant}
\end{figure}

@@

Does the performance of \textit{OneD} on aberrant karyotypes come at the
cost of decreased performance on normal karyotypes? To address this
question, we assembled another data set comprised of mouse B cells and
embryonic stem (ES) cells, both with a normal karyotype. The cell types
were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
the same tests as above were performed.

>>>>

\address{%
$^{1}$Gene Regulation, Stem Cells and Cancer Program, Centre for
Genomic Regulation (CRG), The Barcelona Institute of Science and
Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain
and
$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain
and
$^{3}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
Barcelona, Spain
and
$^{4}$ICREA, Pg. Llu\'is Companys 23, 08010 Barcelona, Spain}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 1~kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Neither of these strategies were designed for cell types with karyotypic
aberrations, most common in cancer. Yet, Hi-C is very sensitive to
aneuploidy, copy number variations and translocations.  Actually, these
aberrations have so much influence on the outcome that they can be used as
signature to re-assemble the target genome \citep{korbel2013genome}. An
additional complication is that karyotypic aberrations are not
experimental biases, so it is unclear whether they should be corrected at
all or be considered part of the biological signal.

@@

\subsection{Model}

@@

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
($\lambda_i'$). The model consists of 8 states that correspond to 1, 2, 3,
4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
emission parameters (a single position parameter for states 1-7, a
position parameter for state `8 or more', a single standard deviation for
all the states and a single degree of freedom for all the states) and 56
transition parameters.

@@

To test the correction of biases, we gathered a set of
published \citep{ledily2014distinct, encode2012integrated, rao20143d,
stadhouders2017transcription, lin2012global, dixon2012topological} and
unpublished Hi-C data of different cell types and organisms
(Supplementary Table~1).

@@

%\begin{landscape}
%
%\begin{table*}
%\label{tab:samples}
%{\begin{tabular}{ccccccc}
%  \hline
%  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
%  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
%  \textbf{Source} \\
%  \hline
%dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
%  DpnII & CRG & T47D, K562 & NA \\
%b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
%  DpnII & CRG & T47D & NA \\
%dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
%  DpnII & CRG & T47D, K562 & NA \\
%HindIII\_T0 & T47D & dilution Hi-C &
%  HindIII & CRG & T47D & SRR1054341 \\
%NcoI\_T0    & T47D & dilution Hi-C &
%  NcoI      & CRG & T47D & SRR1054343 \\
%ENCLB758KFU & T47D & dilution Hi-C &
%  HindIII   & UMass & T47D & ENCLB758KFU \\
%ENCLB183QHG & T47D & dilution Hi-C &
%  HindIII   & UMass & T47D & ENCLB183QHG \\
%HIC069  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & T47D, K562 &    SRR1658693 \\
%HIC070  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & T47D, K562 &    SRR1658694 \\
%HIC071  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
%HIC074  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
%b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%GSM987817 & B cell  & dilution Hi-C &
%  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
%GSM987818 & B cell  & dilution Hi-C &
%  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
%GSM862720 & ES cell & dilution Hi-C &
%  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
%GSM862721 & ES cell & dilution Hi-C &
%  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
%GSM862722 & ES cell & dilution Hi-C &
%  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
%  \hline
%\end{tabular}}{}
%\caption{Hi-C data sets used in this study. Sample ID: Unique sample
%identifier in the respective projects. Cell type: source of the biological
%sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
%digestion. Sequencing core: Laboratory performing the experiment (CRG:
%Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
%Baylor College of Medicine; UCSD: University of California San Diego;
%LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
%sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
%data identifier.}
%\end{table*}
%
%\end{landscape}

@@

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 and mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks}. For more details, see the methods section of
\cite{stadhouders2017transcription}. In addition, the pipeline is
available from the Supplementary Material published by
\cite{quilez2017managing}.

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100~kbp, but we found no major differences at different
resolutions (not shown).

\subsection{Simulations}

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (either B cell or ES cell, and from either
CRG or UCSC). The simulation strategy was the following: First we selected
uniformly at random the amount of copy number break points (from 3 to 10).
We then placed the break points along chromosomes 18 and 19 uniformly at
random. Next, we assigned a copy number (2, 3, 4 or 10 with equal
probability) to each segment delimited by the break points. We computed
the outer product of this simulated copy number profile and multiplied it
element-wise with the original contact matrices of chromosomes 18 and 19.
The resulting matrices were used as input for correction methods. For each
simulation (100 in total), we measured the pairwise reproducibility score
defined by \citet{yan2017hicspector} and computed the average for pairs
from the same cell type minus the average for pairs from different cell
types.

\subsection{Comparison of Hi-C matrices}

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\citet{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

We defined three data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples (see Supplementary
Table~1 for details). Given a set of experiments and a metric, we first
computed all pairwise combinations between experiments and then classified
the comparisons according to the characteristics of each pair (cell type,
protocol, batch and treatment).

@@

\subsection{Bias correction in Hi-C experiments}

The principle of \textit{OneD} is to explicitly model Hi-C biases on a
single variable: the total amount of contacts for each bin of the matrix,
also referred to as the \textit{contact profile}. The reason for this
choice is that the total amount of contacts is approximately proportional
to the local copy number. For instance, a duplicated region in a diploid
genome will show on average a 50\% increase in the number of contacts.
Discontinuities of the amount of contacts thus correspond to changes of
the copy number.

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The brown line represents the raw signal, the blue
line represents the signal after bias correction, the black line
represents the signal after bias and copy number (CN) correction. The long
arm of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
four copies, explaining that the signal is about twice higher than for the
short arm.}
\label{fig:totals}
\end{figure}

Experimental biases affect the contact profile in a continuous but not
necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the number of restriction enzyme sites
in T47D, a breast cancer cell line with an aberrant karyotype.  Four
clouds are visible. Each corresponds to sequences present in one to four
copies. In all of them, the relationship flattens as the number of
restriction sites increases. To capture this relationship, \textit{OneD}
fits a non-linear model between the total amount of contacts and the known
sources of bias (by default the G+C content, the number of restriction
sites and the mappability of the reads).

The experimental biases are estimated genome-wide and each cell of the
Hi-C matrix is then corrected using equation (\ref{eq:xhat}). Note that
the corrected contact profile is still proportional to the copy number.
In cancer cell lines, the contact profile corrected by \textit{OneD} is
highly correlated to the copy number estimated from an external source
(Supplementary Figure~1). Figure~\ref{fig:totals}B shows the corrected
contact profile along chromosome 17 of T47D. \textit{OneD} greatly reduces
the wiggling of the contact profile (dark blue line).

Optionally, \textit{OneD} allows the user to also correct the Hi-C signal
for the copy number. In this case, a hidden Markov model is fitted on the
corrected contact profile in order to infer the local copy number. Each
cell of the Hi-C matrix is then further corrected using equation
(\ref{eq:cnvnorm}), \textit{i.e.} it is divided by the square root of the
inferred copy number. In essence, this approach flatens the contact
profile so that it is no longer proportional to the local copy number. The
light blue line in figure~\ref{fig:totals}B shows the result of this
process. After this correction, the contact profile fluctuates around a
genome-wide constant value.

In what follows, we benchmarked \textit{OneD} and \textit{OneD} with copy
number correction (\textit{OneD+CN}) against \textit{vanilla},
\textit{ICE} \citep{imakaev2012iterative}, the \textit{KR} algorithm
\citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
servant2012hitc}. The first four methods correct biases implicitly,
whereas the fourth method does it explicitly.

We used four metrics to compare Hi-C matrices (see Materials and Methods).
For consistency and concision, the results based on the reproducibility
score of \citet{yan2017hicspector} are shown in the main figures, and the
results based on the other metrics are shown in the Supplementary
Material.

%Given that our model is based on the total number of contacts, we reasoned
%that a preliminary test would be to check if the corrected number of
%contacts per bin reflects the copy number (as measured by an independent
%technique such as array-based copy-number segmentation). We tested the
%validity of this approach against the Catalogue Of Somatic Mutations In
%Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
%shows the Pearson correlation between the corrected number of contacts and
%the copy number estimation for T47D and K562 (a leukemia cell line with an
%aberrant karyotype). Similar results were obtained using Spearman
%correlation (Supplementary Figure 1).  All the methods except
%\textit{OneD} decreased the agreement between the signal and the copy
%number because they partially corrected the discrepancy. In contrast,
%\textit{OneD} enhanced the conformity of the signal with the copy number.
%Not correcting for variable copy number at that stage may seem
%counter-intuitive, but the tests below will show this leads to better
%performance.
%
%
%\begin{figure}
%\centerline{\includegraphics[width=.39\textwidth]
%  {nar_figures/figure_2.eps}}
%\caption{Pearson correlation between the total number of contacts per bin
%and an independent estimation of the copy number (COSMIC). Left panel
%T47D breast cancer cell line, right panel K562 leukemia cell line.}
%\label{fig:copy_number}
%\end{figure}
%

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variance, such as batch effects, laboratory of
origin and protocol variations. Likewise, a good normalization should
increase the contrast between different samples to enhance the
biological differences.

We collected multiple Hi-C data sets obtained from T47D and K562 cells,
both with aberrant karyotypes. The first pool, referred to as the T47D data
set contained six T47D samples and two K562 samples, the second, refered
to as the K562 data set contained eight T47D samples and two T47D samples.
We compared matrices before and after normalization by different methods
(see Materials and Methods, distributions shown in Supplementary Figures~2
and 3). This gave an indication of the impact of a given normalization
method on the biological variation (differences between samples from T47D
and K562) and on the technical variation (differences among samples from
the same cell type). The results are summarized in
Figure~\ref{fig:aberrant}.

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_2.eps}}
\caption{Removing biases from Hi-C on aberrant karyotypes. A and B.
Average changes compared to the raw data on the T47D and K562 data sets.
The bars represent 95\% confidence intervals centered on the mean
difference of the reproducibility score between a given correction method
and the raw data. C and D. ROC curves on the T47D and K562 sets. The areas
under the curve are indicated in the bottom right corner. The color code
is the same as in panels A and B.}
\label{fig:aberrant}
\end{figure}

On the T47D data set, the methods that increased the reproducibility
between identical cell types also increased it between different cell
types (Figures~\ref{fig:aberrant}A), meaning that none of them enhanced
exclusively the biological signal. However, the difference was most marked
for \textit{OneD} and \textit{caICB}. On the K562 data set
(figure~\ref{fig:aberrant}B), both methods increased the reproducibility
between identical cell types but not between different cell types
Figures~\ref{fig:aberrant}B), which is the desired behavior of a
correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
\textit{KR} gave inferior results, while \textit{raw+CN} (copy number
correction alone), \textit{LGF} and \textit{vanilla} were not competitive.
The conclusions were unchanged when using other metrics to compare Hi-C
matrices (Supplementary Figures~4, 5 and 6).

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). \textit{OneD} had the larest area
under the curve in both tests, and the corresponding ROC curve was above
the others throughout. Using other metrics to compare Hi-C matrices
yielded similar results (Supplementary Figures~4, 5 and 6).

Taken together, these results show that \textit{OneD} enhances the
biological variation and reduces the noise on samples with an aberrant
karyotype.

@@

Does the performance of \textit{OneD} on aberrant karyotypes come at the
cost of decreased performance on normal karyotypes? To address this
question, we collected data from mouse B cells and embryonic stem (ES)
cells, both with a normal karyotype. The cell types were pooled in almost
equal proportions (see Supplementary Table~1) and we performe the same
tests as above using the same metrics (distributions shown in
Supplementary Figure~7)

In this test, the reproducibility between different cell types was barely
affected (Figure~\ref{fig:normal}A). All the methods increased the
reproducibility between identical cell types (except one, see below), but
only when the experimental conditions were different. This means that all
the methods were able to remove some technical biases. \textit{ICE} and
\textit{KR} showed the stongest increase in reproducibility, but
\textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
performance. \textit{vanilla} and \textit{LGF} were less competitive, and
\textit{raw+CN} (copy number correction alone) performed poorly, because
it is equivalent to no normalization on diploid cells. Other metrics gave
similar results (Supplementary Figures~8, 9 and 10).
